A Reproducible Data Analysis Workflow with RMarkdown, Git, Make & Docker

### 2019-10-28 rr-flow.Rmd

Here, we combine four software tools, whose interplay can guarantuee full and sustainable computational reproducibility of data analyses and their reporting. There are various tools to enhance reproducibility [@piccoloToolsTechniquesComputational2016], of which we believe four of them are particularly important: literate programming: [@ruleTenSimpleRules2019], version control [@barbaHardRoadReproducibility2016], dependency management @[askrenUsingMakeReproducible2016] and containerization [@clyburne-sherinComputationalReproducibilityContainers2018]. Importantly, only a workflow using all four concepts in unison guarantuee confidence in reproducing a scientific report. Even though there are various implementations of these concepts, we believe that for the R environment <CITATION NEEDED>, the following four implementations are best suited. For literate programming, R Markdown <CITATION NEEDED>, for version control Git <CITATION NEEDED>, for dependency management Make <CITATION NEEDED>, and for container Docker <CITATION NEEDED>. Each of them serves a meta-scientific valuable goal (reproducibility) and increases the productivity of the researcher. Each of these software solutions is highly flexible and powerful, so to master them fully, a considerable amount of practice is necessary. However, for our purposes, it is sufficient to master a minimal valuable subset of functions to ensure reproducibility.

@@

The first step towards reproducibility is to have an R-script or RMarkdown file as primary entry point for the analysis, that runs on a local computer without error. Next, one has to make sure that all files relevant for the analysis can be moved to another computer. To this end, it is essential that all files are within one folder (or, subfolders within it) and all paths are relative to that folder. A robust solution to the problem of making sure that file access does not break across computing platforms are [Rstudio projects](https://r4ds.had.co.nz/workflow-projects.html) and the `here` <CITATION NEEDED> package.

@@

Sometimes external requirements make it impossible that the data can be stored and shared together with the scripts. In most of the cases we have seen, these are either space constraints or privacy considerations. In these cases, of course, unrestricted reproducibility is not guaranteed. If this way of dividing data and scripts into different databases is unavoidable, we recommend to validate all data files using checksums (also called a "hash", e.g., using the functions provided in package `digest` <CITATION NEEDED>) before analyzing them. To do this, a checksum must be created and stored at the time of the original analysis. At the time of reproduction, the current checksum must be compared with the stored checksum.

@@

While our approach was designed to scale well with the complexity of a computing-intense project, we realise that this flexibility may not be straightforward to integrate into a researcher's everyday workflow. There are R packages that implement parts of our workflow and, thus, lower the threshold to adoption when the full flexibility is not needed. The use of R Markdown within a project, tracked with `git` can be simplified with the [workflowR](https://GitHub.com/jdblischak/workflowr) package [@workflowr]. The [drake](https://GitHub.com/ropensci/drake) package [@drake] is directly inspired by `make` but can only handle dependencies within R. The [liftR ](https://cran.r-project.org/web/packages/liftr/vignettes/liftr-intro.html) package [@liftr] and the [holepunch](https://karthik.GitHub.io/holepunch/) package [@holepunch] automatize the use of docker. The former is perfectly compatible with the described workflow and we recommend it to users uncomfortable with command-line use of docker. The later uses [binder](https://mybinder.org) <CITATION NEEDED> to move the analysis to the cloud, so that no local installation of docker is required. `holepunch` is well suited for simple analysis with low computational demands, because of memory and computing time limitations of binder. There are several alternatives to docker that manage dependencies on R packages. [`packrat`](https://rstudio.GitHub.io/packrat/) [@packrat] is a way to freeze package version via local copies of packages in the project, however it does not garuantee a given base R version or system dependencies. Similar approaches are taken by [`jetpack`](https://GitHub.com/ankane/jetpack) [@jetpack], [`miniCRAN`](https://cran.r-project.org/web/packages/miniCRAN/vignettes/miniCRAN-introduction.html) [@miniCRAN] and [`checkpoint`](https://cran.r-project.org/web/packages/checkpoint/vignettes/checkpoint.html) [@checkpoint]. Notable is also the package [`reprex`](https://GitHub.com/tidyverse/reprex) [**repr**oducible **ex**ample, @reprex], but its scope is limited.

@@

However, beside these tools that ease the process of creating a workflow on the researchers side, we see increased interest in changing how research is published and used [@perkelToolkitDataTransparency2018], with *life code* [@perkelPioneeringLivecodeArticle2019] and *continuous integration* [@beaulieu-jonesReproducibilityComputationalWorkflows2017, @yenniDevelopingModernData2018]. These techniques give us a glimpse of a paradigm shift from static to dynamic, interactive and living publications.

>>>>

Here, we combine four software tools, whose interplay can guarantuee full and sustainable computational reproducibility of data analyses and their reporting. There are various tools to enhance reproducibility [@piccoloToolsTechniquesComputational2016], of which we believe four of them are particularly important: literate programming: [@ruleTenSimpleRules2019], version control [@barbaHardRoadReproducibility2016], dependency management [@askrenUsingMakeReproducible2016] and containerization [@clyburne-sherinComputationalReproducibilityContainers2018]. Importantly, only a workflow using all four concepts in unison guarantuee confidence in reproducing a scientific report. Even though there are various implementations of these concepts, we believe that for the R environment [@r2019], the following four implementations are best suited. For literate programming, R Markdown [@rmarkdown2019], for version control Git [@chaconProGit2014], for dependency management Make [@feldmanMakeProgramMaintaining1979], and for container Docker [@merkelDockerLightweightLinux2014]. Each of them serves a meta-scientific valuable goal (reproducibility) and increases the productivity of the researcher. Each of these software solutions is highly flexible and powerful, so to master them fully, a considerable amount of practice is necessary. However, for our purposes, it is sufficient to master a minimal valuable subset of functions to ensure reproducibility.

@@

The first step towards reproducibility is to have an R-script or RMarkdown file as primary entry point for the analysis, that runs on a local computer without error. Next, one has to make sure that all files relevant for the analysis can be moved to another computer. To this end, it is essential that all files are within one folder (or, subfolders within it) and all paths are relative to that folder. A robust solution to the problem of making sure that file access does not break across computing platforms are [Rstudio projects](https://r4ds.had.co.nz/workflow-projects.html) and the `here` [@here2017] package.

@@

Sometimes external requirements make it impossible that the data can be stored and shared together with the scripts. In most of the cases we have seen, these are either space constraints or privacy considerations. In these cases, of course, unrestricted reproducibility is not guaranteed. If this way of dividing data and scripts into different databases is unavoidable, we recommend to validate all data files using checksums (also called a "hash", e.g., using the functions provided in package `digest` [@digest2019]) before analyzing them. To do this, a checksum must be created and stored at the time of the original analysis. At the time of reproduction, the current checksum must be compared with the stored checksum.

@@

While our approach was designed to scale well with the complexity of a computing-intense project, we realise that this flexibility may not be straightforward to integrate into a researcher's everyday workflow. There are R packages that implement parts of our workflow and, thus, lower the threshold to adoption when the full flexibility is not needed. The use of R Markdown within a project, tracked with `git` can be simplified with the [workflowR](https://GitHub.com/jdblischak/workflowr) package [@workflowr]. The [drake](https://GitHub.com/ropensci/drake) package [@drake] is directly inspired by `make` but can only handle dependencies within R. The [liftR ](https://cran.r-project.org/web/packages/liftr/vignettes/liftr-intro.html) package [@liftr] and the [holepunch](https://karthik.GitHub.io/holepunch/) package [@holepunch] automatize the use of docker. The former is perfectly compatible with the described workflow and we recommend it to users uncomfortable with command-line use of docker. The later uses [binder](https://mybinder.org) [@binder2018] to move the analysis to the cloud, so that no local installation of docker is required. `holepunch` is well suited for simple analysis with low computational demands, because of memory and computing time limitations of binder. There are several alternatives to docker that manage dependencies on R packages. [`packrat`](https://rstudio.GitHub.io/packrat/) [@packrat] is a way to freeze package version via local copies of packages in the project, however it does not garuantee a given base R version or system dependencies. Similar approaches are taken by [`jetpack`](https://GitHub.com/ankane/jetpack) [@jetpack], [`miniCRAN`](https://cran.r-project.org/web/packages/miniCRAN/vignettes/miniCRAN-introduction.html) [@miniCRAN] and [`checkpoint`](https://cran.r-project.org/web/packages/checkpoint/vignettes/checkpoint.html) [@checkpoint]. Notable is also the package [`reprex`](https://GitHub.com/tidyverse/reprex) [**repr**oducible **ex**ample, @reprex], but its scope is limited.

@@

However, beside these tools that ease the process of creating a workflow on the researchers side, we see increased interest in changing how research is published and used [@perkelToolkitDataTransparency2018], with *life code* [@perkelPioneeringLivecodeArticle2019] and *continuous integration* [@beaulieu-jonesReproducibilityComputationalWorkflows2017; @yenniDevelopingModernData2018]. These techniques give us a glimpse of a paradigm shift from static to dynamic, interactive and living publications.
